\documentclass[12pt]{article}

\usepackage{mathptmx}
\usepackage{mathpazo}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[version-1-compatibility]{siunitx}
\usepackage{fixltx2e}
\usepackage{multirow}
\usepackage{dsfont}
\newcommand{\inv}{^{-1}}

\topmargin -0.5in %topmargin=1in+\topmargin
\textheight 9in % default is letter % so 11-2=9in
\textwidth 6.5in % and 8.5-2in=6.5in
\oddsidemargin 0in %left margin = % 1in + oddsidemargin
\footskip 1cm % pagenumber to text

\setcounter{secnumdepth}{0}

\begin{document}

\noindent
\textbf{Math 250A, Fall 2015}

\subsection{8/27}

\noindent
A group G acts on a set S:

$G \times S \to S$

$(g, s) \mapsto g \cdot s$

$e \cdot s = s$

$(g g') \cdot s = g \cdot (g' \cdot s)$

\noindent
Alternatively,

$\phi: G \to Perm(S)$ is a homomorphism

$(\phi(g))(s) = g \cdot s$

\noindent
Examples

trivial action: $(\forall g)$ $g \mapsto e_{Perm(S)}$

G acting on self by left/right translation, conjugation

G acting on the set of subgroups of G by conjugation: $g \cdot H = gHg^{-1} = \{ghg^{-1} | h \in H\}$

normal subgroup $N \trianglelefteq G$: all $g \in G$ fix $N$ under conjugation

V vector space over a field K, GL(V) acts on V by $L \cdot v = L(v)$\\

\noindent
The orbit of s, $O(s) := \{g \cdot s | g \in G\}$

constitutes an equivalence relation on S

\noindent
The stabilizer (isotropy group) of $s \in S$, $G_s := \{g \in G |g \cdot s = s\}$

$G_s$ is closed under inverses: $g \in G_s \to g \cdot s = s \to g^{-1}gs = g^{-1}s \to s = g^{-1}s$

\noindent
There exists a natural bijection $\alpha: G/G_s \to O(s)$, $gG_s \mapsto g \cdot s$

well-defined: $g_1G_s = g_2G_s \to \exists g \in G_s, g_1 = g_2g$, $\alpha(g_1G_s) = g_1s = g_2 g s = g_2 s = \alpha(g_2G_s)$

injective: $\alpha(g_1G_s) = g_1 \cdot s = g_2 \cdot s = \alpha(g_2G_s) \to g_2^{-1}g_1 \cdot s = s$, $g_2^{-1}g_1 \in G_s$, so $g_1G_s = g_2G_s$\\

\noindent
Action under conjugation:

the conjugacy classes of a set are the orbits of the action

$O(g) = \{g\} \leftrightarrow g \in Z(G)$ the center of the group

$Z(G) = \{g \in G : xg = gx$ $\forall x \in G\}$

in a permutation group, $\sigma (a_1, a_2, a_3, ... a_k) \sigma^{-1} = (\sigma a_1, \sigma a_2, \sigma a_3, ... \sigma a_k)$

\subsection{9/1}

\noindent
Let $\Sigma$ be a set of representative elements of the orbits of S.

The index of a subgroup H is $(G : H) = \#(G/H)$

For finite G, $(G:H) = \frac{\#G}{\#H}$ ($g \not \in H$, $\exists$ natural bijection $H \to gH$)

$\#S = \sum_{s \in \Sigma}\#O(s) = \sum_s(G:G_s)$

defines a 'mass formula' $\#S = (\sum_s\frac{1}{\#(G_s)})(\#G)$

\noindent
Let G act on a subgroup H by left translation.

$\#H_s = \#H$ and from the above $\#G = (G:H) \cdot \#H$. 

this is a statement of Lagrange's Theorem, $(G:H) = \frac{\#G}{\#H}$.

\noindent
The kernel of the action $K = \bigcap_{s \in S}G_s$, which is just the kernel of $G \xrightarrow{\phi} Perm(S)$.

\noindent
We can relate the stabilizers of points in the same orbit.

Let $s' = gs$.

Assume $x \in G_s$.  

Since $x \in G_s$, $(gxg^{-1})gs = g(xs) = gs$.

Hence $gxg^{-1} \in G_{gs}$, so $gG_sg^{-1} \subset G_{gs}$.

Apply this relation with $g \to g^{-1}$ and $s \to gs$:

Assume $x \in G_{gs}$.

Then $(g\inv xg)(s) = (g\inv)(xgs) = (g\inv gs) = s$.

So $g\inv G_{gs}g \subset G_s \to G_{gs} \subset gG_sg\inv$

\noindent
Thus, $gG_sg\inv = G_{gs} = G_{s'}$.

The stabilizer of $s'=gs$ is a conjugate of the stabilizer of $s$.\\


\noindent
p : prime

\noindent
p-group: a finite group G, $\#G = p^n, n \geq 1$\\

\noindent
``A p-group has a non-trivial center''

Notation: $S^G$ is the set of points in S fixed under the group action. ($gs = s$ $\forall g \in G$)

\noindent
Let G act on itself by conjugation ($S = G$).  Then $S^G = Z(G)$.

For $s \in S (= G)$, $G_s$ is a subgroup, and its order divides the order of the group, $p^n$.

Either $O(s)$ is trivial, and $s \in S^G = Z(G)$, otherwise $\#(O(s)) = p^k$ for $k > 0$

\noindent
$\#S =$ sum of \# of elements in the orbits $\equiv_{mod p} \#$ of orbits of size 1 $= \#(S^G)$.

$\#Z(G) \equiv_{mod p} \#(S^G) \equiv_{mod p} \#S = \#G = p^n \equiv_{mod p} 0$.

$Z(G)$ cannot be 1, since the identity of the group is in the center.

Thus, the order of the center is divisible by p, and must be non-trivial.\\

\noindent
$H \leq G$ a finite group, $(G : H) = p$, the smallest prime dividing \#G $\to H \trianglelefteq G$

Let $S = G\slash H$; $\#(S) = (G : H) = p$, and let G act on S by left translation.

This induces $\varphi: G \to S_P$; recall $\#S_p = p!$

The stabilizer of H, $G_H = \{x \in G | xH = H\}$, hence $G_H = H$.

By inspection, we can see that $G_{gH} = gHg^{-1}$.

Let $K = \bigcap_{g \in G}gHg^{-1}$, the largest normal subgroup contained in H.

For each coset $gH$, $K$ stabilizes that coset, hence $K$ is the kernel of $\varphi$.

By the First Isomorphism Theorem $\varphi(G) \leq S_p$.

$(G : K) = \#(G\slash K) = \#(\varphi(G))$, which divides $\#(S_p) = p!$

Further, since $K \leq H \leq G$, $(G : K) = (G : H)(H : K)$.

Since $(G : K)$ divides $p!$ and $(G : H)$ divides p, $(H : K)$ divides $(p - 1)!$.

But p is the smallest prime dividing $\#G$, so $(H : K)$ = 1, $K = H$ and H is normal.\\

\noindent
A familiar embedding of a group into a larger group; ``Cauchy's Theorem''

$G \hookrightarrow Perm(G)$ by letting G act on itself by left-translation.

Its kernel $K = \{g \in G | gs=s \forall s\} = \{e\}$ (consider $s = e$), so an injection $\to$ an embedding.\\

\noindent
Recall $S_n \subset$ group of $n \times n$ invertible matrices. $\sigma \mapsto M(\sigma)$ a permutation matrix.

Need to be careful in the construction to ensure $M(\sigma\tau)=M(\sigma)M(\tau)$!

E.g. $\sigma = (1 3 2)$ does $M(\sigma)$ have 1 in the 1st column, 3rd row?

Or in the 1st row, 3rd column?  One of these yields $M(\sigma\tau) = M(\tau)M(\sigma)$.

\noindent
G finite of order n; V the vector space of functions $G \xrightarrow{f} \mathds{Z}$; note $V \cong \mathds{Z}^n$

Linear maps $V \to V$ correspond to $n \times n$ matrices over $\mathds{Z}$:  $GL(V) \approx GL(n, \mathds{Z})$.

Similarly, invertible linear maps correspond to $n \times n$ invertible matrices over $\mathds{Z}$.

We can embed G in $GL(n, \mathds{Z})$ by using a left action of G on $GL(n, \mathds{Z}) = \{\phi: V \to V\}$

Can think of this as an action on $\mathds{Z}^n \cong V$, whose permutation group is simply $GL(n, \mathds{Z})$.

Recall that $V = \{f : G \to \mathds{Z}\}$.

This left action takes the form $L_g \mapsto \phi$ where $\phi(f(x)) = f(xg)$

$L_{gg'} = L_{g'} \circ L_{g}$ as desired? Verify for yourself.

Yes: $L_{gg'}(\varphi (x)) = \varphi(xgg') = L_{g'}(\varphi (xg)) = L_{g'} \circ L_g (\varphi(x))$

$g \mapsto L_g$ is a homomorphism $G \to GL(V)$

Using $\mathds{F}_p$ instead of $\mathds{Z}$, get $G \hookrightarrow GL(n, \mathds{F}_p)$, an embedding into a finite group.

\subsection{9/3}

\noindent
Lagrange: If $H \leq G$ then $\#(H)|\#(G)$.

$A_4$ with $n = 6$: a counterexample to the converse.\\

\noindent
If $|G| = p^k \cdot r$, $(p, r) = 1$, a p-Sylow subgroup of G is an $H \leq G$ such that $|H| = p^k$

\noindent
$\mathds{Z}_{12}$ has 2-sylow subgroup $\{0, 3, 6, 9\}$ and 3-sylow subgroup $\{0, 4, 8\}$

\noindent
$D_6$ generated by r, s subject to $rs = sr^{-1}$, $r^6 = e$, $s^2 = e$

$\#(D_6) = 12$ so has 3-sylow subgroup $\{1, r^2, r^4\}$

Also has 2-sylow subgroups $\{1, r^3, s, r^3s\}$, $\{1, r^3, rs, r^4s\}$, $\{1, r^3, r^2s, r^5s\}$\\

\noindent
G = $GL_n(\mathds{F}_p)$, $n \times n$ linear transformations in $\mathds{F}_p$, equal to $Aut(\mathds{F}_p^n)$

\noindent
Approximating the order of $|G|$:  

Asserting linear independence in each vector of an $n \times n$ matrix

$|G| = (p^n - 1)(p^n-p)(p^n-p^2)\cdots(p^n-p^{n-1}) = p^{1 + 2 + 3 + \cdots + n - 1}\cdot r = p^{\frac{n^2 - n}{2}} \cdot r$, $(p, r) = 1$

Consider P the set of $n \times n$ upper triangular matrices with 1's on the diagonal.

Then $|P| = p^{1 + 2 + 3 + \cdots + n - 1} = p^{\frac{n^2 - n}{2}}$, and P is a p-Sylow subgroup.

Will use this fact in the subsequent proof.\\

\noindent
Theorem: (Sylow I) For $|H| = p^k \cdot r$, $(p, r) = 1$, H has a p-Sylow subgroup.

\noindent
Proof Sketch:

Show $\exists G$, $H \leq G$, such that G has a p-Sylow subgroup

Show that if G has a p-Sylow subgroup and $H \leq G$, then H has a p-Sylow subgroup

\noindent
Proof:

Cayley's theorem, can embed $H$ (of order n) in $S_n$ by acting on itself by translation.

Additionally $S_n \leq GL_n(\mathds{F}_p)$ mapping to permutation matrices.

Alternatively, consider $V \cong \mathds{F}_p^n$, the vector space of functions $\varphi : G \to \mathds{F}_p$.

Embed H into $GL(V)$ by the action $g \in H \mapsto$ automorphism taking $\varphi(x)$ to $\varphi(xg)$.

$GL_n(\mathds{F}_p)$ has p-Sylow subgroups.  (upper triangular matrices with 1s on diag)

Let P be a p-Sylow subgroup of $G = GL_n(\mathds{F}_p)$.  Let G act on the cosets of P.

Now, $G_{gP} = gPg^{-1}$. Similarly, when H acts on $G \slash P$, $G_{gP} = (gPg^{-1} \cap H)$

This intersection is a p-group.

Want to choose $g \in G$ such that $gPg^{-1} \cap H$ is a p-Sylow subgroup.

If $(H : (gPg^{-1} \cap H))$ is coprime to p, then $gPg^{-1} \cap H$ is a p-Sylow subgroup.

By Orbit-Stabilizer, $(H : (gPg^{-1} \cap H)) = O(gP)$.

Note this is an orbit of $G/P$ induced by the action of the group H.

Since P is a p-Sylow subgroup of G, $|G\slash P| \not\equiv_{mod p} 0$.

The sum of the orbits is $|G \slash P|$.

Hence there must be some orbit with size coprime to p.

The stabilizer of this orbit $gPg\inv \cap H$ is a p-Sylow subgroup $H_p$.\\

\noindent
Corollary: All p-subgroups of H are contained in a conjugate of P.

Let $J \leq H$ be a p-subgroup.  Then $J \cap gPg\inv$ is a p-Sylow subgroup of $J$ for some $g \in G$.

A p-group can't contain a proper p-Sylow subgroup, so $J \cap gPg\inv = J$ and $J \subset gPg\inv$.\\

\noindent
Corollary: (Sylow II) All p-Sylow groups are conjugate.

Let $H \leq G$ and $P \leq G$ be p-Sylow subgroups.

By the preceding corollary ($G \leq G$, $H \leq G$, $P \leq G$), $H \subset gPg\inv$ for some $g \in G$.

Since $|H| = |P| = |gPg^{-1}|$, $H \cap gPg\inv = H$.\\

\noindent
Corollary: Every p-subgroup of G is contained in a p-Sylow of G.

By the above, each is contained in a conjugate of P, said conjugate being a p-Sylow.\\

\noindent
The p-Sylow subgroups in G are all conjugate, so that:

If P is a p-Sylow of G then $G/N(P) \leftrightarrow$ set of p-Sylows in G.

$N(P)$ the normalizer of P

There are $n_p = (G: N(P))$ p-Sylows in total.\\

\noindent
Lemma: If a finite p-group $\Gamma$ acts on a set X, then $\#(X) \equiv_{mod p} \#(X^{\Gamma})$

($X^{\Gamma}$ the fixed points of X under $\Gamma$).

\noindent
Proof:

Each $\frac{|\Gamma|}{|Stab(x_i)|} \equiv_{mod p} 1$ if $x_i$ fixed, else $\frac{|\Gamma|}{|Stab(x_i)|} \equiv_{mod p} 0$.

Hence $\#X = \sum_i \#Orb(x_i) = \sum_i \frac{|\Gamma|}{|Stab(x_i)|} \equiv_{mod p} \#X^{\Gamma}$.\\

\noindent
Let $Syl_p(G)$ describe the p-Sylow subgroups of G and $n_p$ denote its cardinality.

\noindent
Theorem: (Sylow III) If $|G|= p^k \cdot r$, $k > 0$ then $n_p \equiv_{mod p} 1$.  Further, $n_p | r$.

\noindent
Proof:

Let P act on $Syl_p(G)$ by conjugation.

By the lemma, $\#Syl_p(G) = n_p \equiv_{mod p} (Syl_p(G))^P$.

Suppose Q is fixed under the group action.  Then $pQp\inv = Q$ $\forall p \in P$.

Then $P \leq N(Q)$; similarly $Q \leq N(Q)$.

P, Q are p-Sylow subgroups of N(Q); therefore P, Q are conjugate in N(Q).

However, $Q \trianglelefteq N(Q)$ so that Q is equal to all its conjugates in N(Q), and $P = Q$.

Hence P is the only fixed Sylow-p subgroup so $(Syl_P(G))^P \equiv_{mod p} 1$.

G acts on $Syl_p(G)$ as only one orbit since all p-Sylows in G are conjugate.

$(G : P) = n_p$, $n_p = |G| = p^k \cdot r$, $n_p | p^k \cdot r$, but $n_p \nmid p$, so $n_p | r$.

\subsection{9/8}

\noindent
$P, Q$ p-Sylows and $P \subset N(Q)$ then $P = Q$

reason: $PQ \leq G$ a subgroup of G

$HK$ not necessarily a group, but will be if one normalizes the other ($H \subset N(K)$)\\

\noindent
A simple group is a non-trivial group with no non-trivial proper normal subgroups\\

\noindent
A finite abelian group G is simple $\leftrightarrow$ G is cyclic of prime order

show this\\

\noindent
non-sporadic finite simple groups

$A_n (n \leq 5)$

recall the alternating groups $A_n$ are the even permutations on $\{1, \cdots, n\}$

Lie groups over finite fields, e.g. $\{\pm \begin{pmatrix} 1 & 0 \\ 0 & 1 \\ \end{pmatrix}\} \subset SL(2, \mathds{Z}|p\mathds{Z})$

P = projective; $PSL(2, \mathds{Z}|p\mathds{Z}) = SL(2, \mathds{Z}|p\mathds{Z})$\\

\noindent
Simple groups of order $\leq 60$.

(a) There are no non-abelian simple groups of order $< 60$

(b) If G is simple of order 60, then $G \cong A_5$.

($\#A_n = \frac{n!}{2}$)

\noindent
G simple of order 60.

$H < G$ simple (finite), H proper, $(G: H) = n \geq 2$

G acts on $G\slash H$ by left translation.

The action is transitive (for each pair $xH, yH$, $\exists$ permutation taking one to the other)

Therefore, this action is non-trivial.

$\pi : G \to Perm(G\slash H) = S_n$

$ker(\pi) \neq G$ and is a normal subgroup $\to$ the kernel is trivial.

$\pi: G \hookrightarrow S_n$ and in fact $\pi: G \hookrightarrow A_n$ (if $\#G > 2$)

\noindent
Why? because $G \cap A_n \trianglelefteq G$

If $G \subset S_n$.

Then $G \to S_n/A_n = \{\pm 1\}$ by the sign map, kernel is $G \cap A_n$.

Recall $sgn: S_n \to \{\pm 1\}$ $sgn(\sigma) = (-1)^t$ given t, num of transpositions

$G/(G \cap A_n) \hookrightarrow S_n/A_n = \{\pm 1\}$

$(G : G \cap A_n) = 1$ or $2$.

If G is simple then this cannot be 2 (would be normal subgroup), so =1.

And $G \hookrightarrow A_n$ for that $A_n$.

\noindent
G simple, order 60.

H a proper subgroup of G, index n.  (consider small values of n)

If $n = 2$ then H is normal in G, a contradiction.

(smallest prime dividing the order of a group)

If $n = 3$ or $n = 4$: $G \hookrightarrow A_3, A_4$ but their orders are too small (3, 12)

If $n = 5$: $G \hookrightarrow A_5$ and they are equal in cardinality $\to$ done.

Remaining case: $n = 15$.

What is $n_5$, the number of 5-Sylow subgroups.

$n_5 | 60\slash 5 = 12$, $n_5 = (G : N(P))$ $n_5$ divides the index

Also, $n_5 \equiv_{mod 5} 1$.

Thus $n_5 = 1$ or $n_5 = 6$.

If $n_5 = 1$ then only one 5-Sylow subgroup of G, must be normal.

This is impossible since G is simple.

Then $n_5 = 6$: tells you there are lots of elements of order 5 in G.

There is no overlap (excepting at the identity) between 5-Sylows.

Hence the number of elements of order 5 is $6 \cdot 4 = 24$

Elements of order 5 in $A_5$ are 5-cycles (a b c d e).

Need to take all strings of length 5: 120, and divide out by rotations 5.

Thus we get $120 \slash 5 = 24$ (check).

Consider $n_2$ the number of 2-Sylow subgroups.

Then $n_2$ divides $60/4 = 15$, and $n_2 \neq 1$ because of simplicity.

Also, $n_2 = (G : N(P_2))$, and this can't be 3 since G has no subgroup of index 3.

If $n_2 = 5$ then $N(P_2)$ is the desired index-5 subgroup $\to$ done.

From divisibility $n_2 = 1, 3, 5, 15$.

Elminate 1 by simplicity, 3 since the index is too small, 5 works, consider 15.

\noindent
Considering the situation where there are 15 2-Sylow subgroups (of order 4).

These are groups like the Klein 4-group (no elements of order 4).

There are 2 2-Sylow subgroups P and Q where $P \cap Q$ has order 2.

Prove by counting.

Taking intersection, must be proper else they would be the same.

Hence $P \cap Q$ has order 1 or 2.

If there is utterly no overlap, there are $15 \cdot 3 + 1 = 46$ elt's of 2-Sylows.

And these do not have order 5.  But there are 24 elements of order 5.  Too many.

\noindent
Now we know that some of these 2-Sylow subgroups have non-trivial overlap.

Consider $N(P \cap Q)$ for some such intersection, will be a subgroup of G.

Cannot be all of G, G is simple. (would make $P \cap Q$ normal)

$N(P \cap Q)$ contains P and Q since both are abelian.

Each are normal subgroups of $N(P \cap Q)$, so its order is divisible by 4.

Hence could have order 12, 20, or 60 (divisible by 4, divides 60).

Its index cannot be 1 (G is simple) cannot be 3 ($A_n$ too small), = 5.

QED (\textbf{revisit why}).\\

\noindent
G finite non-trivial.

If G is simple, $\{e\} \subset G$, $G/\{e\}$ simple.

If G is not simple $G \supset G_1 \supset (e)$, $G_1 \trianglelefteq G$, $G_1$, $G/G_1$ smaller than G.

Use principle of strong induction for a full decomposition.

Obtain a successive extension of simple groups.

\noindent
Given $G$, such a tower, let $G_i/G_{i+1} = Q_i$ and consider the multiset $\{Q_0,\cdots , Q_{n-1}\}$.

In multiset, order does not matter, and multiplicity does matter.

\noindent
Jordan-H\"{o}lder Theorem: Each composition yields the same multiset up to isomorphism.

\subsection{9/10}

\noindent
Proposition: Given $G$, $\exists G_0 \supset G_1 \supset G_2 \supset \cdots \supset G_n$, $G = G_0$, $G_{i + 1} \trianglelefteq G_i$, $G_i/G_{i+1}$ simple.

This is a normal tower or composition series; the simple quotients are the constituents.

If it is simple, then the filtration is $G \supset \{e\}$.

If G is not simple, $G \supset N \supset \{e\}$, where $G/N, N$ proper in G.

By strong induction, have filtrations for each.  To conclude, use:\\

$\exists$ natural correspondence between subgroups of $G/N$ and subgroups $H$ of $G$, $N \leq H$

$G \supset L \supset N$, $L/N \subset G/N$

$\pi: G \to G/N$,  $K \subset G/N$, $\to \pi^{-1}(K) \leq G$\\

\noindent
Jordan-H\"{o}lder Theorem:

$G = G_0 \supset G_1 \supset G_2 \supset \cdots \supset G_n$

$G_{i + 1} \trianglelefteq G_i$, $G_i/G_{i + 1} = Q_i$ simple.

The ``multiplicity set'' $\{Q_0, \cdots, Q_{n - 1}\}$ is independent of the filtration.

Where order doesn't count, multiplicity does, and $Q_i$ up to isomorphism.\\

\noindent
Related question: can two different groups have the same reduction?

Yes.  $S_3 \supset A_3 \supset \{e\}.$ Quotients $\mathds{Z}/2\mathds{Z}$ and $\mathds{Z}/3\mathds{Z}$.

Also $\mathds{Z}/6\mathds{Z} \supset 3\mathds{Z}/6\mathds{Z} \supset 6\mathds{Z}/6\mathds{Z}$, same quotients but radically different structure.

``Knowing the building blocks does not confer knowledge of the building.''\\

\noindent
Jordan-H\"{o}lder Theorem: Proof.

Base case $n = 1$, $G \supset \{e\}$, $G/\{e\}$ simple and G simple.

Supposing $G \supset G_1 \supset \cdots \supset G_n \supset \{e\} = G_{n + 1}$ and $G \supset G_1' \supset \cdots \supset G_m' \supset \{e\} = G_{m + 1}'$.

If $G_1' = G_1$, by induction, filtrations of $G/G_1$ and $G/G_1'$ are unique.

If $G_1, G_1'$ distinct, $G_1 \cap G_1' \subsetneq G_1$ and $G_1 \cap G_1' \subsetneq G$

Since $G_1$ and $G_1'$ are normal, $G_1G_1'$ is a subgroup; it is certainly normal

Since $G_1G_1'$ is a proper superset of $G_1$ and $G_1'$ and $G/G_1, G/G_1'$ are simple, $G_1G_1' = G$

Define a map $G_1'/(G_1 \cap G_1') \to G_1G_1'/G_1$; since its kernel is $G_1 \cap G_1'$, it is injective.

This defines $G_1'/(G_1 \cap G_1') \hookrightarrow G/G_1$.  Symmetrically, $G_1/(G_1 \cap G_1') = G/G_1'$.

Have $G_1 \supset \cdots \supset G_n \supset \{e\} = G_{n + 1}$.

Take $G_1 \supset G_1 \cap G_1' = H \supset H_1 \supset H_2 \supset \cdots \supset H_k \supset \{e\}$, a Jordan-H\"{o}lder filtration of $G_1$.

Obtained by induction.

Note $G_1/H = G/G_1'$ is the first quotient of this filtration.

By induction, these two filtrations have the same length.

The constituents of $G_1$ are the constituents of H, with $G_1/H = G/G_1'$ appended.

Constituents: $G/G_1$ + constituents of $G_1$ = $G/G_1$ + $G/G_1'$ + constituents of H.

Have $G \supset G_1' \supset H \supset H_1 \supset \cdots \supset H_k = \{e\}$, same length as $G_1' \supset G_2' \supset \cdots \supset G_m' = \{e\}$.

Have related two different filtrations that have are unrelated, by a common filtration,

which depends on the intersection of these two filtrations.

\subsubsection{Free Groups}

\noindent
S a set, define the free abelian group on $S$, $\mathds{Z}^S = \mathds{Z}\langle S\rangle = \{\sum_{s \in S} n_s \cdot s | n_s \in \mathds{Z}\}$.

Where all but finitely many of the $n_s$ are 0.

$S = \{1, \cdots, n\}$, $\mathds{Z}^S = \mathds{Z}^n = \{(c_1, \cdots, c_n) | c_i \in \mathds{Z}\}$

$$\sum_{i = 0}^\infty n_i x^i = \sum_{i = 0}^\infty n_i \cdot i \in \mathds{Z}\langle S \rangle$$

where $n_i = 0$ for $i >> 0$.

\noindent
``To map $\mathds{Z}\langle X\rangle$ to A in the world of abelian groups is to map S to A in the world of sets.''

$S \to \mathds{Z}\langle S \rangle$ a set map, $s \in S \mapsto 1 \cdot s$.

Given $f: \mathds{Z} \langle S \rangle A$ homomorphism.

And in fact, $F: Hom(\mathds{Z} \langle X\rangle, A) \to Maps(S, A)$, F is a bijection.

\noindent
These elements of the free abelian group are ``formal sums''.

That is, an $f: S \to \mathds{Z}$.

\noindent
Let $f: \mathds{Z}\langle S\rangle \to A$, $f(\sum n_ss) = \sum_{s \in S}n_sf(s)$

\noindent
An abelian group A is free of finite rank if $A \cong \mathds{Z}^n$ for some $n \geq 0$ ($\mathds{Z} = \mathds{Z}\langle \emptyset \rangle = 0$).

Define $rank(A) = n$.  If $\mathds{Z}^m \cong A \cong \mathds{Z}^n$ then n = m.

Why? Take positive integer $> 1$, e.g. 2.  Then $\mathds{Z}^n/2\mathds{Z}^n \cong \mathds{Z}^m/2\mathds{Z}^m$.

LHS has $2^n$ elts and RHS has $2^m$ elts so $n = m$. \\

\noindent
A subgroup of a free abelian group of rank n is a free abelian group of rank $\leq n$.

\noindent
Proof: by induction on n.

$n = 0$: $A = (0) = B$.

$n = 1$: $A = \mathds{Z} \supset B$.  What are the subgroups of $\mathds{Z}$? $(0), (t) = t\mathds{Z}, t \geq 1$.

Proof by division algorithm: $\mathds{Z} \supset B \neq 0$, $t = $smallest positive integer in B.

Division algorithm ensures that all elements are multiples of t.

$B \subset \mathds{Z}^n \xrightarrow{\pi} \mathds{Z}$.

$\pi: (c_1, \cdots, c_n) \mapsto c_n \in \mathds{Z}$.

\noindent
Cases:

(1) $\pi(B) = (0), B \subset \mathds{Z}^{n-1}$, free of rank $\leq n - 1$

(2) $\pi(B) = t\mathds{Z}, t \geq 1$

$B \xrightarrow{\pi|_B} t\mathds{Z} \xrightarrow{surj.} 0$

$ker(\pi)|_B = C$ free of rank $\leq n - 1$.

Choose $b \in B$ such that $\pi(b) = t$.

$C \subset \mathds{Z}^{n-1}: C = ker(\pi)|_B$, free of rank $\leq n - 1$.

$C = B \cap \mathds{Z}^{n - 1}$

$C \subset B$, $\mathds{Z} \cdot b \subset B$

\textbf{Missing (pf in Lang)}

\noindent
Simple linear algebra.

$a_1, \cdots, a_n \in A$ corresponds to a homomorphism $\mathds{Z}^n \to A$, $(c_1, \cdots, c_n) \mapsto \sum_{i = 1}^n c_ia_i$.

These are linearly independent if f is 1-to-1, and these span/generate A if $f$ is onto.

A is finitely generated if A is spanned by $a_1, \cdots, a_n$ for some $n \geq 0$, $a_i \in A$

A is finitely generated iff A is a quotient of $\mathds{Z}^n$ for some n.

\noindent
Corollary: a subgroup of a finitely generated abelian group is again finitely generated.

$\mathds{Z}^n \xrightarrow{f} A$ finitely generated, have $B \subset A$, $f^{-1}(B) \leq \mathds{Z}^n$, and $f^{-1}(B) \cong \mathds{Z}^k$, $k \leq n$.\\

\noindent
A finitely generated, torsion-free.

I.e. given $a \in A$ and $n \cdot a = 0$, $n \geq 1$, then $a = 0$.

Statement: A is free and of finite rank.

\noindent
Proof: Take a finite set of generators S in which take T lin indep and large as possible.

take $T = a_1, \cdots, a_k$ and $S = a_1, \cdots, a_k, \cdots, a_m$

$\sum_{1}^{k + 1} c_ka_k = 0$, $c_{k+1} \neq 0$

$B = span\{a_1, \cdots, a_k\} \cong \mathds{Z}^k$.

$a_{k + 1}, \cdots,a_m$: some multiple lies on B.

$N \geq 1$; $N \cdot A \subset B$.

Th: NA free, $N: A \to NA$ A torsion free.

Multiplication on A by a positive integer is injective.

A is isomorphic to NA by the mutliplication by n, since NA is free, A is free.

\subsection{9/15}

\noindent
Abelian group, finitely generated.

\noindent
Last week:

free group has to do with some correspondence to a $\mathds{Z}^n$

subgroups of free finitely generated abelian groups are free and finitely generated

subgroups of finitely generated abelian groups are finitely generated

finitely generated, torsion free abelian group is a free abelian group

recall torsion free: for all $n \geq 1$, mult by n, $n \cdot A$ is injective

opposite A torsion: for all $a \in A$, $\exists n \geq 1$ such that $n \times a = 0$

\noindent
Example of a torsion abelian group: $\mathds{Q}/\mathds{Z}$

element $p/q \mod{\mathds{Z}}, q \geq 1, p \in \mathds{Z}$; $q \times \frac{p}{q} = 0$ in $\mathds{Q}/\mathds{Z}$

\noindent
finitely generated abelian groups up to isomorphism

A is a direct sum of a free part $\mathds{Z}^r$ and a torsion part (a direct sum of cyclic groups)

\noindent
Direct product of sets $A_i$ indexed by $S$: $$\bigoplus_{i \in S}A_i = \{f: S \to \cup_{i \in S}A_i : f(i) \in A_i\}$$

where for all but finitely many i, $f(i) = 0$

this is equivalent to the direct product when S is finite

\noindent
\textbf{Image 1}: a map from a $\bigoplus_{i \in S}A_i$ to B is determined by the mappings from the $A_i$

The direct sum is a coproduct.

\noindent
\textbf{Image 2}: a map into a $\prod_{i \in S}A_i$ is determined by the mappings into the $A_i$

The direct product is a product (in the categorical sense).

\noindent
$S$ countably infinite, $A_i = \mathds{Z}/2\mathds{Z}$

$\bigoplus_{i \in s}A_i$ is countable, but $\prod_{i \in S}A_i$ is not

\noindent
Categories: products, coproducts, morphisms

$Mor(?, B) = \prod Mor(A_i, B)$ $? =$ co-product

The coproduct of sets is disjoint union.

\noindent
Abelian group A and subgroups X and Y

we have inclusions from each into A

$X \times Y = X \oplus Y \xrightarrow{h} A$, $(x, y) \mapsto x + y$

$h$ is injective if every $a \in A$ is of the form $x + y$

$h$ is one-to-one $\leftrightarrow$ you can't write $x + y = x' + y'$ unless $x = x'$, $y = y'$

If true, say A is the direct sum of its submodules X and Y.

\noindent
Suppose A, $X \subset A$, $A/X$ is free (f.g. free): then X has a complement Y in A, $A \cong X \oplus A/X$

$A \xrightarrow{\pi} A/X$

$Y \subset A$, $\pi|_Y$ is an isom $Y \to A/X$.

$\pi|_Y$ inj $\leftrightarrow Y \cap X = (0).$

$\pi|_Y$ surjective: given $a + X \in A/X$ we can find $y \in Y$ s.t. $y + X = a + X$

$x = y \cdot a \in X$

$a = y \cdot x, x\in X, y\in Y$

$A/X$ free, say $\cong \mathds{Z}^r$

To map $A/X$ to $A$ is to choose images in $A$ of the generators of $A/X$ corresponding to the unit vectors of $\mathds{Z}^r$.

There is a unique homomorphism $s: A/X \to A$ so that $s(q_i) = a_i$ for $i = 1, \cdots, r$

$(\pi \cdot s)(q_i) = \pi(a_i) = q_i$

$\pi \circ s = id_{A/X}$

$Y =$ image of $S \subset A$.

$\pi|_Y$ surjective.  $\pi(s(q)) = q$ for all $q \in A/X$

$\pi|_Y$ is 1-1.  $/pi(s(q_0)) = 0$ but $s(q_0) = q_0$ so equals 0.

\noindent
A a finitely generated abelian group

$X = A_{tors} = \{a \in A | na = 0$ for some $n \geq 1\}$.

X f.g., tors $\to$ X finite abelian group.

$A/X$ torsion free, f.g. $\to$ $A$ free $\approx \mathds{Z}^r$

\noindent
$A \approx \mathds{Z}^r \oplus A_{tors}$.  $A_{tors} = ???$

it is a finite abelian group, let $B = A_{tors}$

p prime, $B_p = \{b \in B | p^t \cdot b = 0$ for some $t \geq 0$\}.

$B_P \subset B$.

$\bigoplus_p B_p \xrightarrow{\iota} B$

Proposition: $\iota$ is an isomorphism. (formal proof in Lang's book)

\noindent
Proof essence:

suppose $60 \cdot b = 0$, $60 = 4 \cdot 3 \cdot 5 = 12 \cdot 5$

$(12, 5) = 1$

$1 = r5 + s12 = 25 - 24$

$b = r \cdot 5 \cdot b + s \cdot 12 \cdot b$

$12x = 0$, $5y = 0$

Every element can be written as a sum of terms killed by a power of a prime

\noindent
$A = \mathds{Z}^r \oplus (\bigoplus_p B_p)$

\noindent
$\mathds{Z}^n \approx F \xrightarrow{\varphi} A$ $A$ finitely generated (by n elements)

$Ker(\varphi) = X \subset F$.

? understand A ! understand X inside F.

\noindent
Elementary division theorem

There exists a basis of $F \approx \mathds{Z}^n$ s.t. ... $X = \bigoplus_{i \leq r}0 \oplus a_1\mathds{Z} \oplus a_2\mathds{Z} \oplus \cdots \oplus a_{n - r}\mathds{Z}$, $a_i \geq 1$

$X \subset \mathds{Z}^n$

$a_1 | a_2 | a_3 | \cdots | a_{n - r}$, increasing multiplicatively

$A = F/X = \mathds{Z}^r \oplus \mathds{Z}/a_1\mathds{Z} \oplus \mathds{Z}/a_2\mathds{Z} \oplus \cdots$, $a_i | a_{i + 1}$

A a finite abelian group $\to$ A is a direct sum of cyclic groups

\noindent
p prime, $\#A = p^4 = a_1a_2a_3\cdots$

A is direct sum of cyclic groups of p-power order.

$A \approx \mathds{Z}|p^i \oplus \mathds{Z}|p^j \oplus \mathds{Z}|p^k \oplus \mathds{Z}|p^l$ at most

$i \leq j \leq k \leq l$, $i + j + k + l = 4$, $i, j, k, l, \geq 1$

\subsection{9/17}

\noindent
A arbitrary finitely generated group that we want to understand

Pick some generators $g_1, \cdots, g_n$

Get a map from $Y = \mathds{Z}^n$ to $A$, has some kernel

Considering $A = Y/X$, and how $X$ lies in $Y$ gives indication of structure of $A$

Can think of $X$, $Y$, as lattices

\noindent
Theorem: $Y \cong \mathds{Z}^n$ exists $v_1, \cdots, v_n$ basis of Y

such that in that basis $X = a_1\mathds{Z} \oplus a_2\mathds{Z} \oplus \cdots \oplus a_m\mathds{Z} \oplus 0 \oplus 0 \oplus \cdots \oplus 0.$

$a_i \geq 1$, $a_1 | a_2, a_2 | a_3, \cdots a_{m - 1}|a_m$.

\noindent
Example: $A = \mathds{Z}/2\mathds{Z} \oplus \mathds{Z}/3\mathds{Z}$

$Y = \mathds{Z} \oplus \mathds{Z}$

$Y \supset X = 2\mathds{Z} \oplus 3\mathds{Z}$

Not at all true that the integers divide each other.

Puzzle.  Not the case as in the theorem.

Need to prove to self that in some new basis, $Y = \mathds{Z} \oplus \mathds{Z}$, 

and $X = \mathds{Z} \oplus 6/\mathds{Z}$, $Y/X = \mathds{Z}/6\mathds{Z}$.

$a_1 = 1$, and $a_2 = 6$.\\

\noindent
$X \subset \mathds{Z}^n$.  Ask whether $X = (0)$ the zero submodule.  If so, simple.  So can assume nonzero.

Consider linear forms, homomorphisms $\mathds{Z}^n \to \mathds{Z}$.

For each $\lambda$ have $\lambda(X) \subset \mathds{Z}$.  e.g., $\lambda(X) = 3\mathds{Z}$.  Some $\lambda$s are nonzero since $X$ is nonzero.

Choose $\lambda$ so that $\lambda(X)$ is maximal.

Example: $X = 2\mathds{Z} \oplus 3\mathds{Z}$.  The first coordinate fn yields $2\mathds{Z}$, 

the second coordinate fn yields $3\mathds{Z}$.

But with $\lambda(u, v) = v - u$ we can get all of $\mathds{Z}$.

possible to get $\lambda$s yielding images $2\mathds{Z}$, $3\mathds{Z}$, but not to get $\lambda$, $\lambda(X)$ containing both?

In any case, take a maximal $\lambda$, fix that $\lambda$.

$\lambda(X) = a\mathds{Z}$ maximal

Pick $x \in X$ so that $\lambda(x) = a$.

\noindent
Claim: $\mu(x) = b$ is divisible by a for all $\mu \in Hom(\mathds{Z}^n, \mathds{Z})$

$gcd(a, b) = g = ra + sb$

$\tau := r\lambda + s\mu$, $\tau(x) = g$

Now $\tau(X) \supset \mathds{Z}g \supset \mathds{Z}a$

So $\tau(x) = \lambda(x)$, $\mathds{Z}g = \mathds{Z}a$

$a | b$ for this reason of maximality

\noindent
``Executive session''

R a commutative ring

R-module: M

1) abelian group

2) endowed with a scalar multiplication $r \in R$, $m \in M$, $rm \in M$

same as a vector space definition except $R$ is not assumed to be a field

\noindent
The context in which this elementary divisor theorem works.

A a finitely generated abelian group replaced by a finitely generated R-module

And there are 2 conditions on R.

R is an integral domain: $rs = 0 \to r = 0$ or $s = 0$

Ideals of R are principal $M \subset R \to M = R\cdot a$

\noindent
Digression: motivation.  Killer example.

K a field, and R = K[t]. (very much like $\mathds{Z}$, can do Euclidean division by remainders)

Have V and action of $K[t]$: (action of $K$ and action of $t$)

V + action of K $\to$ K-vector space

Action of t: $T: V \to V$ multiplication by t, $v \mapsto t \cdot v$, $T(v) = t \cdot v$

Conversely, can form the corresponding polynomial in the linear transformation

Principal Ideal domain.  Element of smallest degree, Euclidean algorithm.

Suppose we have an R-module V.  This is a K-vector space V with action of t

Multiplication by t gives a linear operator $T: V \to V$ (t commutes with K)

\noindent
Remark: if V is of finite dimension over K, then it is finitely generated as a K-module

In particular, it's finitely generated over the ring $R = K[t]$

\noindent
A an abelian group.  If A is torsion, we are especially interested.

\noindent
Suppose we start with a linear operator on a finite-dimension vector space.

There is a characteristic polynomial $h$ such that $h(T) = 0$.

Cayley-Hamilton theorem.

$h(t) \in R = K[t]$.  So $h(t) \cdot v = 0$.

V is a torsion module because $h(t)$ annihilates V.

\noindent
Summary of what we have so far:

$0 \neq X \subset Y = \mathds{Z}^n$, $\lambda: Y \to \mathds{Z}$, $\lambda(X)$ is maximal among $\mu(X)$s, $\lambda(X) = a\mathds{Z}$.

Have shown that $a = \lambda(x)$, then $\mu(x)$ is divisible by a for all $\mu$.

Take $\mu$ to be the $i^{th}$ coordinate function, $x = (x_1, \cdots, x_n) \in \mathds{Z}^n$, $a|x_i$ for all $i = 1, \cdots, n$, 

$x = a \cdot y$, $y \in \mathds{Z}^n$, $\lambda(y) = \lambda(x)/a = 1$

\noindent
Think of Y: contains two submodules (subgroups)

$Y \supset ker(\lambda)$, $Y \supset \mathds{Z} \cdot y$.

Claim: $Y = ker(\lambda) \oplus \mathds{Z}y$

1) each $z \in Y$ is: e.g. $(z - \lambda(z) \cdot y) + \lambda(z)y$

2) if $my$ is in $ker(\lambda)$ then $0 = \lambda(my) = m\lambda(y) = m$ so $m = 0$, $my = 0$, intersection is 0

\noindent
The corresponding statement for X is that $X = (\ker(\lambda|_X)) \oplus \mathds{Z}_X$

Kind of obvious that the intersection is 0.

Each component is a submodule of the corresp. one in Y.

$z \in X$, $\lambda(z) = m\lambda(x) = ma\lambda(y)$.

$z = z - \lambda(z)y + \lambda(z)y$

$\lambda(z)y = m\cdot a\cdot y = mx$

$(z - \lambda(z)y) \in ker(\lambda) \cap X = ker(\lambda|_X)$

\noindent
$\mathds{Z}^n = Y = ker(\lambda) \oplus \mathds{Z}y$

$Y \supset X = ker(\lambda|_X) \oplus \mathds{Z}ay$

Apply inductively to portion of lower rank, having pulled off $\mathds{Z}a$

\noindent
$X = a_1\mathds{Z} \oplus a_2\mathds{Z} \oplus \cdots \oplus a_m\mathds{Z} \oplus 0 \cdots 0 \subset Y = \mathds{Z} \oplus \mathds{Z} \oplus \mathds{Z} \oplus \cdots \oplus \mathds{Z}$

need to have some kind of divisibility among these $a$, need to be explained

$a_1|a_2, \cdots$

\noindent
$Y = \mathds{Z} \oplus Y'$ and $X = a\mathds{Z} + X'$, working rightward

start thinking of various linear maps $\lambda': Y' \to \mathds{Z}$, and how they restrict to X

taking a maximal one, etc., etc.

need to understand somehow that if we take this $\lambda'(X') = a'\mathds{Z}$

we want $a | a'$, meaning $a'\mathds{Z} \subset a\mathds{Z}$, do this with some greatest common divisor argument

\noindent
Introduce $g = gcd(a, a')$ which we want to be $a$, write in form $ra + sa'$

Need to find some interesting linear map from Y to Z

Have a map $Y' \xrightarrow{\lambda'} \mathds{Z}$ and $\mathds{Z} \to \mathds{Z}$ the identity

Both of these are linear maps that give linear maps $Y \to \mathds{Z}$.

Choose $x' \in X'$ so that $\lambda'(x') = a'$

Have $(a, 0)$ in X so that the second linear map (just taking the first coordinate)...

...applied to $(a, 0)$ gives a

\noindent
Take $Y = \mathds{Z} \oplus Y'$

$\mathds{Z} \oplus Y' \xrightarrow{f} \mathds{Z}$

$\mathds{Z} \oplus Y' \to Y' \to Y' \xrightarrow{\lambda'} \mathds{Z}$, the composition of which call $g$

$Y = \mathds{Z} \oplus Y' \ni (a, x') \in X$

$f(a, x') = a$

$g(a, x') = \lambda(x') = a'$

$(rf + sg)(a, x') = G$, $rf + sg = \mu$

\noindent
$\mu(X) \supset \mathds{Z} \cdot G \supset \mathds{Z}a$

Maximality $\to G = a$.

Tells us that $a$ really divides $a'$ by maximality.

\noindent
The Y and the X really divide off into two separate worlds.

$Y = \mathds{Z} \oplus Y'$ and $X = a\mathds{Z} \oplus X'$

The world which we have already considered, and the trailing-off world of $Y'$ and $X'$

New map $\mu$ defined on all of $Y$ and $X$, by leaving the first coordinate alone.

\noindent
Go back to the original example of the 2 and the 3.  $Y = \mathds{Z} \oplus \mathds{Z} \supset X = 2\mathds{Z} \oplus 3\mathds{Z}$

$\lambda(u, v) = v - u$

$x = (2, 3), \lambda(x) = 1$

$a = 1, \lambda(X) = \mathds{Z}$, need to see how that line splits off in $\mathds{Z}$ and in X.

$Y = \mathds{Z}\cdot y \oplus ker(\lambda)$

$y = x/a = x$, $ker(\lambda) = \{(u, v) : u = v\} = \mathds{Z} \cdot (1, 1)$

$Y = \mathds{Z} \cdot (2, 3) \oplus \mathds{Z} \cdot (1, 1) = \mathds{Z}^2$

$X = \mathds{Z} \cdot (2, 3) \oplus \mathds{Z} \cdot (1, 1) \cap (2\mathds{Z} \oplus 3\mathds{Z})$

so $X = \mathds{Z} \cdot (2, 3) \oplus 6 \cdot \mathds{Z}(1, 1)$

$Y/X = \mathds{Z}/\mathds{Z} \oplus \mathds{Z}/6\mathds{Z} = \mathds{Z}/6\mathds{Z}$.

\subsection{9/22}

\noindent
Rings R, A ($=$ 'anneau')

definition: whether or not $1 \in R$ can vary

Lang: $1 \in R$, Hungerford: $1 \not \in R$

In the former, $2\mathds{Z}$ is not a ring, in the latter, it is

\noindent
Ring:

under +, an abelian group with distinguished element 0

under $\cdot$, associative (not necessarily commutative) with distinguished element 1

distributive laws $(x + y)z = \cdots$ and $z(x + y) = zx + zy$

\noindent
Integral domain:

\noindent
Field: under $\cdot$, commutative, and non-zero elements have inverses\\

\noindent
Examples

\noindent
For A an abelian group, the ring of endomorphisms.

$R = End(A) = Hom(A, A)$, $(f + g)(a) = f(a) + g(a)$, $fg = f \circ g$

If $A = \mathds{Z}^n$ End(A) can be viewed as a ring of matrices

\noindent
$\mathds{Q}, \mathds{R}, \mathds{C}, \mathds{Z}/p\mathds{Z}, \mathds{Q}(i) = \{a + bi | a, b \in \mathds{Q}\}$ are fields.

\noindent
The ``skew field'' of Hamilton quaternions over $\mathds{R}, \mathds{Q}$, $a + bi + cj + dk$ ($= (\frac{a - bi - cj - dk}{a^2 + b^2 + c^2 + d^2})\inv$)

\noindent
Group G (written multiplicatively), $\mathds{Z}[G] = \mathds{Z}\langle G \rangle$ the free abelian group on G

elements $\sum n_g \cdot g, n_g \in \mathds{Z}$ the sum finite

can multiply

$$(\sum n_g \cdot g)(\sum m_h \cdot h) = \sum_{x \in G} (\sum_{g, h, gh= x}n_g m_h)x$$

The term $c_x = \sum_g n_g m_{g^{-1}x}$ represents a convolution product\\

\noindent
Ring Homomorphism

a homomorphism of abelian groups respecting multiplication

$\varphi(xy) = \varphi(x)\varphi(y)$

$\varphi(1) \neq 1$ is possible

$ker(\varphi) = \{r \in R | \varphi(r) = 0\}$ is an ideal: $x \in R, r \in ker(\varphi) \to xr, rx \in ker(\varphi)$\\

\noindent
Ideals

additive subgroup with ideal property:

$xI \subset I$ left-sided , $Ix \subset I$ right-sided, both $\to$ 2-sided (bilateral)

exact analogues of normal subgroups

\noindent
two-sided ideal: well-defined quotient multiplication

$(r + I) \cdot (s + I) := rs + I$

$(r + I)(s + I) = r(s + i) + I = rs + ri + I$ and similarly

$(r + I)(s + I) = (r + i)s + I = rs + is + I$

therefore ideals are kernels of ring homomorphisms

\noindent
Principal Ideal $(a)$ is the minimal ideal containing $a$

is all multiples of $a$ in $R$: $I = Ra$, said to be generated by $a$

\noindent
Ideal generated by a subset X is the intersection of all ideals containing X

if $X = \{a_1, \cdots, a_t\}$, is written $(a_1, \cdots, a_t)$

\noindent
Prime ideal $P \subset R$

proper

if $rs \in P$ then $r \in P$ or $s \in P$\\

\noindent
Examples

\noindent
$\mathds{Z}$ has as ideals the additive subgroups $a\mathds{Z}$, $a \geq 0 = (a)$; if $a=0$ or $a$ is prime, $(a)$ is prime

\noindent
$R = K[x]$ where $K$ is a field: by Euclidean division, all ideals are principal

\noindent
$R = K[x, y]$

$R \xrightarrow{\varphi} K$, $f(x, y) \mapsto f(0, 0) \in K$, $ker(\varphi) = \{$polynomials with constant term of 0$\}$

this is \textit{not} principal

elements $0 + ax + by + cx^2 + \cdots$\\

\noindent
$\varphi: R \to S$ a ring homomorphism and $P$ a prime ideal of $S$

$\varphi^{-1}(P)$ is a prime ideal of $R$

\noindent
Proof:

Let $x, y \in R$ and suppose $xy \in \varphi^{-1}(P) = P'$

then $\varphi(x)\varphi(y) = \varphi(xy) \in P \to \varphi(x) \in P$ or $\varphi(y) \in P$\\

\noindent
Corollary: $\varphi: R \to S$ a non-trivial homomorphism of rings and $(0)$ prime in $S$

the kernel of $\varphi$ is prime\\

\noindent
S is called an integral domain if 

$(0) \neq S$

if $xy = 0$ then $x = 0$ or $y = 0$

\noindent
Proposition: $P \subset R$ is a prime ideal $\leftrightarrow R/P$ is an integral domain\\

\noindent
Maximal ideal $M \subset R$

$M \neq R$

$M \subset M' \to M = M'$ or $M' = R$

\noindent
Proposition: M is maximal $\leftrightarrow$ $R/M$ is a field

\noindent
Example: $\mathds{Z} \supset a\mathds{Z}$ maximal $\leftrightarrow a$ is prime

\noindent
Corollary: Maximal ideals are prime

Pf: Fields are integral domains.

\subsection{9/29}

\noindent
(Charlie)

\noindent
$A$ a ring, $I$ an ideal in $A$

have a correspondence between ideals $J$ of $A$ containing $I$ and the ideals of $A/I$

$\pi: A \to A/I$ and $\pi(J) = J/I$ an ideal of $A/I$

for $K$ ideal of $A/I$, $\pi^{-1}(K)$ is an ideal of $ A$\\

\noindent
$A$ a ring, its group of units $A^* = \{u \in A | \exists v \in A, uv = 1\}$

$(\mathds{Z}[i])^* = \{1, -1, i, -i\} \cong \mathds{Z}/4\mathds{Z}$

$(\mathds{R}[x])^* = \mathds{R}^*$

$(\mathds{Z}[\sqrt{5}])^* \ni 1, -1, 2 + \sqrt{5}, 2 - \sqrt{5}$\\

\noindent
$A$ a field $\leftrightarrow$ $A^* = A - \{0\}$ and $A \neq \{0\}$

$\{0\}$ is a maximal ideal\\

\noindent
Every proper ideal of $A$ is contained in a maximal ideal.

Proof by Zorn's Lemma.\\

\noindent
Chinese Remainder Theorem

a ring A with ideals $I_1, \cdots, I_k$, $k \geq 2$

the ideals coprime: that is, $I_i + I_j = A$.

then there exists a surjective map $A \to A/I_1 \times \cdots \times A/I_k$

\noindent
example

$r\mathds{Z} + s\mathds{Z} = gcd(r, s)\mathds{Z}$

$(r\mathds{Z})(s\mathds{Z}) = r\cdot s\mathds{Z}$

$r\mathds{Z} \cap s\mathds{Z} = lcm(r, s)\mathds{Z}$

$(lcm)(gcd) = rs$

\noindent
for two: $IJ$, $A/(IJ) \leftrightarrow (A/I) \times (A/J)$

($IJ = I \cap J$)

\noindent
Proof:

Assume $I, J \subset A$, $I + J = A$

$A \to A/I \times A/J$

let $x + y = 1$

$x \to (0, 1)$ and $y \to (1, 0)$, $cx + dy \to (c, d)$\\

\noindent
Quotient Fields

e.g. $\mathds{Z} \to \mathds{Q}$

$A$ an integral domain and $S$ a ``multiplicative subset'' of A

$1 \in S$, $x, y \in S \to xy \in S$

$S\inv A = $ equivalence class

\subsection{10/1}

\noindent
Principal Ideal Domain: $\forall$ ideals $I$, $I = (a)$

\noindent
Noetherian ring

every ideal is finitely generated $I = (a_1, \cdots, a_m) = \{\sum_{i=1}^mr_ia_i | r_i \in A\}$

\noindent
$I_1 \subset I_2 \subset I_3 \subset \cdots$ increasing chain of ideals in A

becomes stable: $\exists N \geq 1$ so that $I_n = I_N$ for all $n \geq N$

e.g. in $\mathds{Z}$, have $(2^{100}) \subset (2^{99}) \subset \cdots$ (arbitrarily long chains exist, but all terminate)\\

\noindent
the following are equivalent

(1) each ideal is finitely generated

(2) chains become stable

(3) every non-empty set of ideals of A contains a maximal element.

\noindent
(1) implies (2)

given, $I_1 \subset I_2 \subset \cdots$ take $I = \bigcup_{i = 1}^\infty I_i$

$I$ finitely generated, each $a_i$ needs to be in some $I$

eventually all of them are in some $I_N$, so $I \subset I_N$ and we are done

\noindent
(2) implies (3)

$S$ some set of ideals, $I_1 \in S$.  If $I_1$ not maximal, $I_1 \subset I_2$, $I_2 \in S$, iterate to construct a chain

by (2), becomes stable; $I_N$ is maximal

\noindent
(3) implies (1)

$I$ an ideal, $a_0 \in I$, $I \neq (a_0)$, $\exists a_1$, $a_0 \subsetneq a_1$

iterate $\to$ ascending sequence: has a maximal element $(a_0, \cdots, a_r) = I$\\

\noindent
irreducible elements of A cannot be factored

$a \in A$, not a unit and $\neq 0$

if $a = bc$ then $b$ is a unit or $c$ is a unit

\noindent
$(0) \subset (a) \subset A$; if A is a principal ideal domain, $(a)$ is maximal

$I \supset (a) = (b) \subset A \to a \in (b)$, $a = bc$

$b$ a unit $\to I = A$, $c$ a unit $\to I = (a)$\\

\noindent
Principal ideal domain $A$, $t \in A$, $t \neq 0$, $t$ not a unit

Proposition: $t$ can be written as a product of irreducible elements

\noindent
Proof:

Let $S=$ the set of (principal) ideals $(t)$ for which the proposition is false

If nonempty, has maximal element $(m)$; if $(m) \subsetneq (m')$, $(m')$ can be factored

$m$ irreducible else $m = m'm''$ where $m'$, $m''$ not units

$(m) \subsetneq (m')$, $(m) \subsetneq (m'')$, hence neither are in $S$

\noindent
Proof works for Noetherian rings generally\\

\noindent
prime elements of A

$a \neq 0$, not a unit, $a$ prime $\leftrightarrow (a)$ is prime

if $a | bc$ then $a | b$ or $a | c$

\noindent
Primes are irreducible:

if $a$ is prime and $a = bc$ then $a|b$ or $a|c$

if $a|b$ then $b$ is a multiple of $a$ and $a$ is a multiple of $b$

so $a \sim b$: $b = u \cdot a$ and $a = u^{-1} \cdot b$, differ by a unit\\

\noindent
irreducible elements might not be prime

$A = \mathds{Z}[\sqrt{-3}] = \{a + b\sqrt{-3} | a, b \in \mathds{Z}\}$

$2 \cdot 2 = (1 + \sqrt{-3})(1 - \sqrt{-3})$

$2$ is irreducible and not prime, $2 | 4$ but doesn't divide either on the right side

exists norm $N: z \mapsto z\overline{z}$

$a + b\sqrt{-3} \mapsto a^2 + 3b^2$

$2$ is irreducible

$2 = \alpha\beta$, $N(2) = N(\alpha)N(\beta)$, $4 = N(\alpha)N(\beta)$

but norms can never be $2$ so one of these must be a unit ($N = 1$ implies $\pm 1$)\\

\noindent
In a principal ideal domain, irreducible elements are prime

$(a)$ irreducible $\to (a)$ maximal $\to (a)$ is prime $\to a$ is prime\\

\noindent
Unique factorization domain: every $a \neq 0$, unit has a factorization as a prod of irreducibles

this is unique up to reordering and transformation by units

$a \sim b$, a and b are associated, if $a = b \cdot u$ and $b = a \cdot u^{-1}$ for some unit $u$

\noindent
Theorem: PIDs are UFDs

PID: $a = \pi_1 \cdots \pi_n = \sigma_1 \cdots \sigma_m$

$\sigma_m$ prime so $\sigma_m$ divides some $\pi_i$

can assume $\sigma_m | \pi_n$, $\phi_n = \sigma_m \cdot c$, $c$ unit

proceed by induction on indices, end

\noindent
A PID $a, b \in A$, $(a, b) = \{ax + by | x, y \in A\} = (g)$ since principal

$g = gcd(a, b)$: $(g) = (a, b) \ni a, b$

$a$ and $b$ are multiples of $g$, $g$ divides $a$, $b$

\noindent
t can't be factored as a product of irreducibles, $(t)$ is maximal in this property

\noindent
if t irreducible $t = t$; impossible

if $t$ not irreducible, $t = r \cdot s$, $r, s$ non-units

$(t) \subsetneq (r)$ $(t) \subsetneq (s)$

\noindent
$A = \mathds{Z}[\cdots (7)^{\frac{1}{2^N}} \cdots]$

7 is not a unit in $A$

\noindent
Lemma: every element of A is ``integral''

it satisfies an equation (monic polynomial) $x^n + c_{n-1}x^{n-1} + \cdots + c_0 = 0$

monic: first coefficient $= 1$

$c_i \in \mathds{Z}$

integral ring

\noindent
$1/7$ satisfies no such polynomial

\noindent
7 can be factored on and on $n(7^{1/n})$; not a Noetherian ring

\subsection{10/6}

\noindent
A-Modules (left modules)

M = abelian group with an action of scalar multiplication of A (= ring)

(same axioms as for an A-vector space except that $A \neq $field)

\noindent
$End(M) = Hom(M, M)$

$M = \mathds{Z}^n$, $End(M) = M(n, \mathds{Z})$

action of A on M: a homomorphism of rings $A \xrightarrow{\varphi} End(M)$

$\varphi(a) \in End(M)$, $\varphi(a): M \to M)$, $(\varphi(a))(m) := a \cdot m$

$f, g \in End(M)$: $fg = f \circ g$

\noindent
Diversion: Fresh water (Chicago) algebra: $a \in A, m \in M, m^a, (m^{ab}) = (m^a)^b$

instead of $a \cdot m)$ or $a(m)$

\noindent
Module properties

$\varphi(ab) = \varphi(a)\varphi(b)$

$(ab) \cdot m = a \cdot (b \cdot m)$

$a \cdot (m + m') = a \cdot m + a \cdot m'$

$\varphi(a) \in End(M)$

$(a + b) \cdot m = a \cdot m + b \cdot m$

$\varphi(a + b) = \varphi(a) + \varphi(b)$\\

\noindent
Examples:

$A =$ field: an A-module is an A-vector space

Th: (uses choice) every A-vector space has a basis $\leftrightarrow$ all A-modules are free

\noindent
M free on the set of generators $\{x_i\}_{i \in I}$

if every $m \in M$ is uniquely a finite A-linear combination of the $x_i$

\noindent
For I, the free A-module on the set I

$\{\sum_{i \in I}a_ix_i | a_i \in A$ all but finitely many are 0$\}$

could also notate $\{\sum_{i \in I}a_ii | a_i \in A$ all but finitely many are 0$\}$, just indexed by I

\noindent
Direct sums $\{M_i\}_{i \in I}$, $\oplus_{i \in I}M_i$

set of tuples indexed by $I$, with the $i^{th}$ entry in $M_i$, all but finitely many entries are 0

$a \cdot (\cdots m_i \cdots)_{i \in I} = (\cdots am_i \cdots)_{i \in I}$

\noindent
Homomorphisms of A-modules $M, N$

$M \xrightarrow{h} N$, conditions of linearity $h(x + y) = h(x) + h(y)$, $h(a \cdot x) = a h(x)$

$A =$ field: linear map

$Hom_A(M, N)$ is an A-module

\noindent
A map from a direct sum to a module uniquely determined by action on the summands

$$M \hookrightarrow \oplus_{j \in I}M_j \xrightarrow{h} N$$ $$M_i \xrightarrow{h_i} N$$ $$Hom_A(\oplus M_i, N) \xrightarrow{\alpha} \prod_{i \in I} Hom_A(M_i, N), h \mapsto (\cdots, h_i, \cdots)$$

$\alpha$ is a bijection

To map a free module to N is to choose the images of each of the generators

Unconstrained: can choose arbitrarily the images of the generators\\

\noindent
Examples

\noindent
$A = \mathds{Z}$, $M =$ ab grp, $\mathds{Z} \to End(M)$, $1 \mapsto \varphi(1) = id$, $2 \mapsto id + id$, $-1 \mapsto -id$

\noindent
$A = A, I \subset A$ left ideal, $I = $A-module, $a \cdot i = ai \in I$

\noindent
ring hom $A \to A'$, $M = A'$-module, $A \to A' \xrightarrow{\varphi} End(M)$, A'-modules $\mapsto$ A-modules

\noindent
M = $\mathds{Z}$-module, $n \geq 1$, $M^n = \oplus_{i = 1}^n M$

$A = M(n, \mathds{Z})$ acts on $M^n$ by left matrix multiplication

could replace $\mathds{Z}$ by some ring R, new construction

An exercise: A-modules $\leftrightarrow$ abelian groups, leftwards, $M \mapsto M^n$, rightwards, ?

Morita equivalence

\noindent
Exact sequence $X \xrightarrow{h} Y \xrightarrow{g} Z$; $Im(h) = Ker(g)$ (implies $g \circ h = 0$, but even stronger)

can make these as long as we like $\cdots X_1 \xrightarrow{f_1} X_2 \xrightarrow{f_2} \cdots$

exact if exact at each place $X_i$, i.e. $Ker(f_{i + 1}) = Im(f_i)$ for all $i$\\

\noindent
Examples

\noindent
$Y \xrightarrow{g} Z \xrightarrow{0} 0$, exact.  $g$ is surjective (epimorphism)

\noindent
$0 \to X \xrightarrow{h} Y$, exact.  $h$ is injective (monomorphism)

\noindent
$0 \to X \xrightarrow{h} Y \xrightarrow{g} Z \to 0$ is called a short exact sequence.  $Y/h(X) \cong Z$

\noindent
$X \xrightarrow{h} Y$, $0 \to Ker(h) \to X \xrightarrow{h} Im(h) \to 0$, exact, $X/Ker(h) \cong Im(h)$

\noindent
$0 \to Im(h) \to Y \to Coker(h) \to 0$

\noindent
$0 \hookrightarrow Ker(h) \hookrightarrow X \xrightarrow{h} Y \to Y/Im(h) = Coker(h) \to 0$\\

\noindent
$N \to X \to Y$ $N \to Y$, $0 \to X \to Y \to Z \to 0$ exact.  $Hom_A(N, X) \to Hom_A(N, Y)$

use a functor, get a $0 \to Hom(N, X) \to Hom(N, Y) \to Hom(N, Z) \to 0$

have exactness at $Hom(N, X)$, $Hom(N, Y)$

what about exactness at $Hom(N, Z)$?

equivalent statement: every homomorphism $N \to Z$ lifts to a homomorphism $N \to Y$

the entering map not necessarily surjective

e.g. $A = \mathds{Z}, X = 2\mathds{Z}. Y = \mathds{Z}$ and $Z = Y/X = \mathds{Z}/2\mathds{Z}$, $N = \mathds{Z}/2\mathds{Z}$, lift does not exist

go from left to right using functor/construction $Hom_A(N, \cdot)$

this functor/construction is ``left exact'' but not ``right exact/fully exact''

the class of modules with full exactness are the projective modules

\subsection{10/8}

\noindent
(Tal)

\noindent
A a ring and M, N modules

$Hom_A(M, N)$ is an abelian group (addition pointwise)

if $A$ is commutative, then it is an $A$-module

\noindent
$0 \to X \xrightarrow{f} Y \xrightarrow{g} Z \to 0$ an exact sequence

$0 \to Hom_A(N, X) \to Hom_A(N, Y) \to Hom_A(N, Z) \to 0$ 

this sequence is left exact and exact at the center

\noindent
surjectivity of the map $Hom_A(N, Y) \to Hom_A(N, Z)$

$Y \twoheadrightarrow Z$ via $G$, $h: N \to Z$

does $H$ exists such that $g \circ H = h$?

the same question, rephrased:

suppose we map $Hom_A(N, Y) \to Hom_A(N, Z)$, taking $H$ to $h$

is this map surjective?

\noindent
an example of a case where it does not lift

take $h > 1$

$\mathds{Z} \to \mathds{Z}/h\mathds{Z}$ surjective

identity $\mathds{Z}/h\mathds{Z}$, look for map $\mathds{Z}/h\mathds{Z} \to \mathds{Z}$

map doesn't exist, no lifting\\

\noindent
(1) Suppose $y \xrightarrow{g} Z$ is surjective.

If $Hom_A(N, Y) \xrightarrow{g_*} Hom_A(N, Z)$ is also surjective, we say $N$ is projective

an equivalent statement: the functor $Hom_A(N, \cdot)$ is right exact

another equivalent statement: for all $g, h$ there is a lifting $H$

$g: y \twoheadrightarrow Z$, $h: N \to Z$, $H: N \to y$

\noindent
(2) given a sequence $0 \to X \xrightarrow{f} Y \xrightarrow{g} N \to 0$

if $\exists s$ such that $g \circ s = id_N$ we say that the sequence splits

all exact sequences split (misreading notes?)

given $y \xrightarrow{g} N \to 0$, we can find $s$ such that $g \circ s = id$

\noindent
(1) implies (2)

if $N$ is projective, then the exact sequence $0 \to X \xrightarrow{f} Y \xrightarrow{g} N \to 0$ splits

take $h = id \to Z = N$

$y \xrightarrow{g} N$, $id: N \to N$, $N \to Y$

\noindent
(3) the module $N$ is a direct summand of a free module:

$\exists M$ such that $N \oplus M \cong F$ where $F = A\langle S \rangle$

\noindent
(2) implies (3)

choose a set of generators of $N$, call it $S$.  You can have $N \subset S$

induces $A\langle S \rangle \to N$ surjective

we have $f: N \to A\langle S\rangle$ by hypothesis

so $A\langle S\rangle = Ker(g) \oplus f(N)$

\noindent
(3) implies (1)

$F = M \oplus N$ where $F$ is free

want to show that if $g: Y \twoheadrightarrow Z$ then $Hom_A(N, Y) \xrightarrow{g_*} Hom_A(N, Z)$ is surjective

$F = A\langle S \rangle$ same as $Hom_A(F, X) = Maps(S, X)$

$S$ generates $F$, so a map $F \to X$ is determined by $S$

$Hom_A(F, Y) = Maps(S, Y)$ and $Hom_A(F, Z) = Maps(S, Z)$

$Maps(S, Y) \to Maps(S, Z)$ obviously surjects, $Hom_A(F, y) \to Hom_A(F, Z)$ surjects

$s \to z \in Z$ and $Y \ni y \to z \in Z$

$Hom_A(M \oplus N, y) = Hom_A(M, Y) \times Hom_A(N, Y)$

have surjective $Hom_A(M \oplus N, y) \xrightarrow{\sigma} Hom_A(M \oplus N, Z)$

$Hom_A(M, Y) \times Hom_A(N, Y) \xrightarrow{(g_*, g_*)} Hom_A(M, Z) \times Hom_Z(N, Z)$

since $\sigma$ surjective, $(g_*, g_*)$ surjective and $g_*$ surjective

Thus $Hom_A(N, Y) \xrightarrow{g_*} Hom_A(N, Z)$ surjective.\\

\noindent
Diagram

$Hom(M \oplus N, y) \to Hom(N, Y)$ by $h \mapsto h \circ i$

$Hom(M \oplus N, y) \xrightarrow{g_*} Hom(M \oplus N, Z)$

$Hom(N, Y) \to Hom(N, Z)$

$Hom(M \oplus N, Z) \to Hom(N, Z)$

$g \circ h \mapsto (g \circ h) \circ i = g \circ (h \circ i)$

\noindent
Diagram

$N \xrightarrow{h \circ i} Y$

$N \hookrightarrow M \oplus N$

$M \oplus N \xrightarrow{h} Y$\\

\noindent
Examples: free modules are projective.

Any free module is a summand of another module that generates (word unsure) a free module

If A is a field, all A-modules are free.

\noindent
$A = K \oplus K = \{(a, b) | a, b \in K\}$, $K$ a field

$F = A = N \oplus M$ where $N = \{(a, 0)| a \in K\}$ and $M = \{(0, b) | b \in K\}$

Projective, but not free over A.

Suppose $N \cong A\langle S\rangle$, basis over $A \leftrightarrow S$

$N \cong A^n$, $dim_kN = 2h = 1$ (not sure about these figures)

$n > 1$, $A = M(n, k)$, $F = A$

$M \in A$, $x \in F$

$M_1x$ matrix (not sure if right), $x = (c_1 \cdots c_n)$ n columns

$M \circ X = (M_{c_1}, \cdots, M_{c_n})$

$F = K^n \oplus \cdots \oplus K^n$

$K^n$ projective, not free.

\noindent
example (justification left for homework)

$k$ a number field; that is, contains $Q$ ($\mathds{Q}$?) and $dim_Qk < \infty$

let $\alpha \in \mathds{C}$ and $\alpha$ algebraic

$k = span(1, \alpha, \alpha^2, \cdots, \alpha^{n - 1})$

$k$ field

\noindent
Diagram

$A \subset k$, $\mathds{Z} \subset \mathds{Q}$, A assoc with $\mathds{Z}$, k with $\mathds{Q}$

\noindent
$A = \{\beta \in k| \beta$ satisfies a monic polynomial with integer coefficients$\}$\\

\noindent
Theorem: the ring $A$ is a Dedekind domain.

\noindent
Definition of a Dedekind domain

$I$ an ideal in $A$, then there exists $J \subset A$ (is it an ideal?) such that$$IJ = \{\sum_{r = 1}^t x_r y_r | x_r \in I, y_r \in J, t \geq 0\}$$ is principal

What is J?

Define $I\inv = \{y \in k | yI \subset A\} \supset A$

$I\inv$ is an $A$-submodule of $k$.

$II\inv \subset A$, in fact $II\inv = A$

\noindent
Theorem: If $I$ is a nonzero ideal in a Dedekind Domain $A$, $I$ is a projective.

\subsection{10/13}

\noindent
Category

notation e.g. $\mathcal{A}$, (sets)

objects, and $Mor(A, B)$ a set of morphisms from the object $A$ to the object $B$

axioms: every object has an identity morphism

composition of which preserves morphisms, etc.

define isomorphisms in terms of the existence of inverses

in some categories, bijections are not isomorphisms

\noindent
Examples

for $A$ a ring, the category of $A$-modules

morphisms of which are the $A$-linear homomorphisms $X \to Y$, $Hom_A(X, Y)$

pointed sets $(X, x)$ in which a morphism is an $f:(X, x) \to (Y, y)$, $f(x) = y$

$A$-modules $X, Y, Z, W$ and fixed $X$ take as objects pairs $(Y, f)$ where $f: Y \to X$

we have created a new category relative to $X$

category of partially ordered sets, whose morphisms are isotone maps\\

\noindent
Functor

takes objects to objects, and also morphisms to morphisms

\noindent
Diagram($F$ is the functor, $f$ is a morphism, $A, A'$ are objects)

$A \xrightarrow{F} F(A)$

$A \xrightarrow{f} A'$

$A \xrightarrow{F} F(A')$

$F(A) \xrightarrow{Ff} F(A')$

since the arrows go in the same direction, this desribes a covariant functor

if, say, $F(A') \xrightarrow{Ff} F(A)$, this would be a contravariant functor

\noindent
Examples

forgetful functors from for instance (groups) $\to$ (sets) or ($A$-modules $\to$ (abelian groups)

Fix $X$. Functor from $A \in Ob(A) \to Mor(X, A)$ (morphisms in the category of sets)

or, contravariantly $A \in Ob(A) \to Mor(A, X)$

in $A$-modules, fix $X$ and take $N$ to $N \oplus X$

or from (sets) to (abelian groups) using the free group construction\\

\noindent
Representable Functors

covariant $\mathcal{A} \xrightarrow{F}$ (sets)

Fix $X$.  By the hom-functor $h_X$, $A \mapsto Mor(X, A)$.

Given an $F$, can it be written as a hom-functor?

That is, for some $X$, is $F \cong h_X$?

Those which can be are said to be represented by $X$ (not a complete definition)

\noindent
Fully defining a representable functor $F$

we need an $X \in \mathcal{A}$ and a $u \in F(X)$ such that for all $A$ have a bijection $$Mor(X, A) \to F(A)$$

corresponding an $h: X \to A$ to a morphism $h_*: F(X) \to F(A)$

the lower-star signifies a covariant (push-forward)

a contravariant (pull-back) would be represented by an upper-star

can associate $h \in Mor(X, A)$ with $h(u)$ and this $h \mapsto h(u)$ is a bijection

epithet: to give an element of $F(A)$ is to give a map $X \to A$\\

\noindent
Example of a Representable Functors

Fix a set S, let $\mathcal{A}$ be the category of abelian groups

Take $G \mapsto F(G) = Maps(S, G)$

Want an abelian group $X$ such that $Maps(S, G) \cong Hom(X, G)$

Take $X$ to be the free abelian group on $S$

The universal element $u$ is the set map taking $s$ to $1 \cdot s$

\noindent
Diagram:

$X = \mathds{Z}\langle S \rangle \xrightarrow{h} G$

$S \xrightarrow{u} \mathds{Z}\langle S \rangle$

$S \xrightarrow{h_*(u)} G$

\noindent
a set map in the category of sets is given by a group map from the free group\\

\noindent
Another example

From $\mathcal{A}$ the category of abelian groups to sets

Fix $M, N \in \mathcal{A}$.  Define the functor $A \mapsto Hom(M, A) \times Hom(N, A)$

to give a pair of maps $M \to A$, $N \to A$ is to give a map from the direct sum to $A$

Take $X = M \oplus N$ and $u \in F(X) = Hom(M, X) \times Hom(N, X)$

$u$ is a universal pair of inclusions

to give a map of the direct sum is to give a map of the first and a map of the second\\

\noindent
The uniqueness of $(X, u)$

if they represent the same functor, they are isomorphic in a canonical sense

no choice involved in the formulation of isomorphism

say $(X, u)$ and $(X', u')$ represent the functor $F$

then $Mor(X, A) \ni h \mapsto h_*(u) \in F(A)$ and $Mor(X', A) \ni h' \mapsto h_*'(u') \in F(A)$

taking the particular cases when $A = X'$, $A = X$

\textit{not totally sure if the next two lines are totally right}

\textit{I remember he said in class that these are ``the same'' h as those in the above line}

there is a bijection $Mor(X, X') \to F(X')$; so for some $h \in Mor(X, X')$, $h(u) = u'$

there is a bijection $Mor(X', X) \to F(X)$ so for some $h' \in Mor(X', X)$, $h'(u') = u$

the representing property of $X$ and $X'$ gives two morphisms

their compositions are the identity on $X$ and the identity on $X'$ (why?)\\

\noindent
Tensor Products

can be defined on noncommutative rings

one must be a left-module and the other a right-module

will be defined on a commutative ring $\mathcal{A}$ for simplicity

$A$-modules $X, Y, Z, M, N, T$

bilinear maps $Bil(X \times Y, Z)$: linear in each variable

i.e. $Bil(X \times Y, Z) = Hom_A(X, Hom_A(Y, Z))$

\noindent
two examples of bilinear maps

$X, Y = k^n$ for $k$ a field, $f(x, y) = det(x | y | c_1 | \cdots | c_{n - 2}) \in k$

$x \in X$, $Y = Hom_A(X, A) \ni \varphi$ a linear form, $(x, \varphi) \mapsto \varphi(x)$\\

\noindent
Define a functor whose representing element is $T$ a tensor product.

Fix $X, Y$ in the category of $A$-modules and have $F: Z \mapsto Bil(X \times Y, Z)$

$F(Z) = Mor(T, Z) = Hom_A(T, Z)$

$u \in F(T)$ gives the universal bilinear map $u: X \times Y \to T$

A homomorphism $T \to Z$ gives a bilinear map $X \times Y \to Z$

i.e. there is a set bijection $Bil(X \times Y, Z) \leftrightarrow Mor(T, Z)$

How one constructs such a $T$: next lecuture.

\noindent
$T$ has uniqueness property by canonical isomorphism.

We do some amount of work to show this construction is possible

Then we can abstract away this work because of the universality of $T$

(last line unsure; maybe ask about it again)

\subsection{10/15}

\noindent
$T$ a universal object: is $u \in F(T)$

$\forall z \in F(A)$ exists unique $h \in Mor(T, A)$ such that $z = h_*u$

\noindent
Representable Functor: example

$\mathcal{A} =$ (rings), $F: \mathcal{A} \to$ (sets) 'forgetful', $F(A) = A$

want a ring $T$ and $u \in T$ such that $h \in Mor(T, A)$ corresponds to $h(u) \in A$

Take $T = \mathds{Z}[x]$, $u = x$

$Hom(\mathds{Z}[x], A)$ determined by image of $x$

so mapping $\mathds{Z}[x]$ to $A$ $\leftrightarrow$ choosing elt of $A$\\

\noindent
A non-representable functor: example

$\mathcal{A} =$ (rings) and $F(A) = \{a \in A | a = b^2, b \in A\}$

representability corresponds to automorphisms!

\noindent
non-representability

$A = \mathds{Z}[x]$, $A \xrightarrow{\alpha} A$ an involution ($\alpha^2 = Id$) defined $f \mapsto (x \mapsto f(-x))$

Assume $F$ representable by $(T, u)$; say $u = v^2$.

In definition, take $A = \mathds{Z}[x]$, $z \in F(A)$, $z = x^2$.

Exists a unique homomorphism $h: T \to \mathds{Z}[x]$ such that $h(u) = x^2$

Here we're not using the notation $h_*$ since $h_*$ is the restriction of $h$ to the squares

$(\alpha h)(u) = (-x)^2 = x^2$

$h(u) = h(v^2) = x^2 \to (h(v))^2 = x^2 \to h(v) = x$ or $h(v) = -x$

$((\alpha h)(v))^2 = x^2 \to (\alpha h)(v) = -x$ when $h(v) = x$ and $(\alpha h)(v) = x$ when $h(v) = -x$

$\alpha h$ and $h$ differ on $v$ so they are distinct, violating uniqueness of $h$\\

\noindent
Tensors again

Commutative ring $A$; $\mathcal{A} =$ ($A$-modules); $M, N$ fixed $A$-modules

$F(X) = Bil(M \times N, X)$

Theorem (construction): $F$ representable by $T = M \otimes_A N$ and $u: M \times N \to M \otimes_A N$

all bilinear maps $M \times N \xrightarrow{b} X$:

unique homomorphism of $A$-modules $T \xrightarrow{h} X$, $h \circ u = b$

$(m, n) \mapsto m \otimes n$ pure tensors

Fact: all elements of $T$ are sums of pure tensors\\

\noindent
Example from linear algebra

$A = K$ a field, $V = M$, $W = N$

If $V$ has dimension $m$ we have a basis $v_1, \cdots, v_m$, similarly $w_1, \cdots, w_n$

$V \otimes_K W$ has dimension $mn$, basis $v_i \otimes w_j$

A similar expression can be found if we have rings and $v_i, w_j$ finitely generate them\\

\noindent
Tensor identities

\noindent
$A \otimes _A N$: consider $Bil(A \times N, X) \leftrightarrow Hom_A(A, Hom_A(N, X)) = Hom_A(N, X)$

$T = N$, $u \in Bil(A \times N, N): (a, n) \mapsto an$

Conclusion: $A \otimes_A N = N$ with universal bilinear map $u$

\noindent
$(M_1 \oplus M_2) \otimes_A N \cong (M_1 \otimes_A N) \oplus (M_2 \otimes_A N)$

$Bil((M_1 \oplus M_2) \otimes N, X) = Hom_A(M_1 \oplus M_2, Hom_A(N, X))$

$= Hom_A(M_1, Hom_A(N, X)) \times Hom_A(M_2, Hom_A(N, X))$

$= Bil(M_1 \times N, X) \times Bil(M_2 \times N, X) = Hom_A(M_1 \otimes_A N, X) \times Hom_A(M_2 \otimes_A N, X)$

$=Hom_A((M_1 \otimes_A N) \oplus (M_2 \otimes_A N), X)$

tensor products commute with direct sums\\

\noindent
Tensor Product: the construction

$Bil(M \times N, X) \subset Maps(M \times N, X) = Hom_A(A\langle M \times N \rangle, X)$

pass to a quotient of $A\langle M \times N \rangle$ satisfying conditions

conditions include $h((am, n)) = h(a(m, n))$

take $T$ to be the largest quotient of $A\langle M \times N\rangle$ in which $(am, n) = a(m, n)$

this is the quotient by the submodule generated by $(am, n) - a(m, n)$, etc.

$u: M \times N \to T$ taking $(m, n)$ to the image of $(m, n)$ in the free module $:= m \otimes n$\\

\noindent
$A = K$, $V = K \otimes \cdots \otimes K$ (m times), $W = K^n$

$V \otimes_K W = (K \otimes \cdots \otimes K) \otimes (K \oplus \cdots \oplus K)$ = big sum of $K \otimes K$s, $K \otimes_K K = K$

\noindent
$M \otimes_A N \cong N \otimes_A M$, $m \otimes n \mapsto n \otimes n$

Consider bilinear maps $M \times N \to N \otimes_A M$, $(m, n) \mapsto n \otimes m$

universality: bilinear maps f factor through unique map $h: m \otimes n \mapsto n \otimes m$

$b: (m, n) \mapsto m \otimes n$ 

\noindent
Tensors and sequences

$X \xrightarrow{f} Y$, $N$ $\to$ $X \otimes N \to Y \otimes N$ defined $f \otimes id$

Proposition: If $f$ is onto, then $f \otimes id$ is surjective.

Need it to make all pure tensors in $Y \otimes N$, since these are generators

This is simple

\noindent
A short exact sequence $0 \to X \xrightarrow{f} Y \xrightarrow{g} Z \to 0$, ($f$ injective, $g$ onto, $Im(f) = Ker(g)$)

$X \otimes N \xrightarrow{f \otimes id} Y \otimes N \xrightarrow{g \otimes id} Z \otimes N \to 0$

call $F = f\otimes id$ and $G = g \otimes id$

$G \circ F = 0$, $Im(F) \subset Ker(G)$

proposition: this new sequence is exact; i.e. $Ker(G) \subset Im(F)$

thus the tensor product is right-exact\\

\noindent
$\mathds{Z}/a\mathds{Z} \otimes \mathds{Z}/b\mathds{Z} = \mathds{Z}/g\mathds{Z}$ where $g = gcd(a, b)$

\noindent
Example

$A = \mathds{Z}$.  $0 \to \mathds{Z} \to \mathds{Z} \to \mathds{Z}/2\mathds{Z} \to 0$, first map is multiplication by two

$N = \mathds{Z}/2\mathds{Z}$.

Get an exact sequence of $\mathds{Z}/2\mathds{Z}$s.

\subsection{11/3}

\noindent
$A =$ UFD (e.g. $\mathds{Z}$, ...), will show that $A[x]$ is UFD.

$p$ prime of A irreducible element (up to mult by units)

$A \subset K$ quotient field

$a \in A$, $a \neq 0$: $a = p^n$ (elt not div by $p$), $n \geq 0$.

Def: $n = ord_pa$.

$t \in K^*$, $t = \frac{a}{b}$, $ord_pt := ord_pa - ord_pb$, $ord: K^* \to \mathds{Z}$

$ord_p(0) = \infty$

for $f(x) \in K[x]$, $ord_pf(x) =$...

$\infty$ if $f(x) = 0$

otherwise the minimum of the $ord_p$ of the coefficients of $f(x)$

$K = \mathds{Q}$

$f(x) = (\frac{1}{3}x^3 + \frac{1}{2}x^2 + x + 7)10^3$

$ord_pf(x) = $

for $p \geq 5$, $0$

for $p = 2$, 2 (division of $1/2 \cdot 10^3$)

for $p = -3$, -1 (division of $1/3 \cdot 10^3$)

for $p = 5$, 3

\noindent
Content cont(f) = $\prod_p p^{ord_{p}f(x)}$

take the positive primes or the negatives, but not both

e.g. $cont(f(x)) = 2^2 \cdot 3^{-1} \cdot 5^3$\\

\noindent
$\frac{f(x)}{cont(f)} \in A[x]$, has content 1

\noindent
$cont(c \cdot f(x)) = c \cdot cont(f)$

\noindent
Gauss's Lemma

$cont(f \cdot g) = cont(f) \cdot cont(g)$

proof: replace $f$ by $\frac{f}{cont(f)}$, $g$ similarly

$fg \in A[x]$, need $cont(fg) = 1$

$(p)$ prime ideal $A \to A/(p)$ = integral domain contained in its quotient field

$h(x) \in A[x] \mapsto \bar{h}(x) \in A/(p) [x]$

Want $p \not | fg$ (all p)

Hyp: $p \not |f$ and $p \not |g$

$\bar{f}, \bar{g} \neq 0 \to \bar{fg} \neq 0$ since in integral domain\\

\noindent
Cor: If a poly in $A[x]$ factors non-trivially in $K[x]$, it factors in $A[x]$ as a product of polynomials of positive degree

factors non-trivially: $f(x) = g(x)h(x)$ where degrees of $g, h$ are positive

replace $f$ by dividing out its content, now has content 1

$f(x) = g(x)h(x)$

certainly $\frac{f(x)}{cont(g)cont(h)}$ factors nontrivially in $A[x]$

has factors $\frac{g(x)}{cont(g)}$ and $\frac{h(x)}{cont(h)}$

think through: why it is that a poly with content 1 is in $A[x]$

has to do with unique factorization and ord = 0 for all primes

\noindent
Observation: $A$ UFD $\to$ $A[x]$ UFD with primes:

primes of $A$

irreducible polys in $K[x]$ scaled to have content 1\\

\noindent
$f(x) \in A[x]$, not constant

$f(x) = p_1(x) \cdots p_t(x)$ in $K[x]$

this is equal to $\frac{p_1(x)}{cont(p_1)} \cdots \frac{p_t(x)}{cont(p_t)} \cdot cont(f)$

have to check if there are two different ways to write poly, essentially the same\\

\noindent
$A[x_1, \cdots, x_n]$ is a UFD

this equal to $(A[x_1, \cdots, x_{n - 1}])[x_n]$

$x^2 + y^2 - 1$

$K[x, y]$ x is irred $(x)$ is prime

$(x) \subset (x, y) \subset K[x, y]$

$(x, y)$ maximal

thing is no longer a PID

$0 \to (x, y) \to K[x, y] \to K$

$f(x, y) \mapsto f(0, 0)$

$\mathds{Z}[x]$ PID UFD 

prime ideal $(7) \subset (7, x)$ (not maximal)

easy to find PID that are not UFD

going to rings that are not dimension 1

\noindent
Exercise: if $A$ is a Dedekind domain (Dedekind ring), $A$ is a PID $\leftrightarrow$ $A$ is a UFD\\

\noindent
Eisenstein's criterion

$f(x) \in A[x]$, $p$ prime, $f(x)$ non-constant

say $f(x)$ has degree $n$

$p \not | a_n$, $p | a_k$ for $k \neq n$, $p^2 \not | a_0$

conclusion: $f(x)$ irreducible in $K[x]$

\noindent
Proving Eisenstein's criterion (proving irreducibility by p-adic methods)

Assume reducible; $f(x) = g(x)h(x)$ give $g$ coefficients up to $b_m$, $h$ coefficients up to $c_d$

and $g, h \in A[x]$ by Gauss's Lemma

now, $m, d \geq 1$ and $m + d = n$

either $p |b_0$ and $p \not | c_0$ or vice versa, assume wlog former

$f(x): p \not | b_m$, $b_mc_d = a_n$ not divisible by $p$\\

$t \leq m$, $m < n$

$a_t = b_tc_0 + b_{t - 1}c_1 + \cdots b_0c_t$

by hypothesis, all of these in sum divisible by $p$

by hypothesis, $a_t$ is divisible by $p$

contradiction, $b_t$ and $c_0$ are not divisible by $p$\\

\noindent
Example: over $\mathds{Q}$. $p$ a prime number

$x^p - 1$ with roots $e^{\frac{2\pi i}{p}}$ etc.

Prop: $\sum_{k = 1}^{p - 1}x^k$ is irreducible

$f(x) = g(x + 1) = 1 + (x + 1) + \cdots + (x + 1)^{p - 1} = x^{p - 1} + a_{p - 2}x^{p - 2} + \cdots + a_1x + p$

highest coefficient is 1 and the constant coefficient div by $p$ but not $p^2$

Lemma: intermediate coefficients $a_1, a_2, \cdots a_{p - 2}$ all div by $p$

take $f(x) mod p$ this is equal to $x^{p-1}$

then $g(x + 1) mod P = x^{p - 1}$ and $g(x) mod p$ = $(x - 1)^{p - 1}$

$g(x) = \frac{x^{p} - 1}{x - 1}$

$f(x) = \frac{(x + 1)^p - 1}{(x + 1) - 1} = \frac{(x + 1)^p - 1}{x}$

$(x + 1)^p - 1 = xf(x)$ now think mod p

mod p, $(x + 1)^p = x^p + 1$

mod p $(x + 1)^p - 1 = x^p$ and mod p $xf(x) = x \bar{f}(x)$

$\mathds{Z}/p\mathds{Z}[x]$

leading to $\bar{f}(x) = x^{p - 1}$

\end{document}
